The ChEMBL database as Linked Open Data

### 2013-04-14 article.tex
Applied Antonty's linguistic fixes

The ChEMBL version used in this paper is ChEMBL 13, which was released on 29 February 2012.
The SQL data dump provided by ChEMBL was inserted into a local MySQL database. A set of SQL queries 
were then executed to generate RDF triples using a set of PHP scripts. The scripts use SQL to
query the database and output triples using print statements. Custom PHP scripts have been used
as a light-weight solution to provide a Linked Data API via a Apache webserver, as well as to
create data files with triples. These scripts are Open Source 
and are available from the source code hosting service, GitHub~\cite{ChEMBLRDFGitHub}. This process 
has been outlined in a best practices note by the W3C Health Care and Life Sciences Interest
Group~\cite{Marshall2012}. This process is similar to its original description.

However, the previous ChEMBL-RDF conversion used custom predicates and classes to represent RDF triples, defining
an ad-hoc ontology~\cite{Willighagen2011}. This implicit ontology reflects the concepts expressed
in the relational database, and is not formally defined. The current version, instead, uses community-proposed
ontologies, making the resulting RDF triples more interoperable with RDF provided by other life sciences databases. 
The new triple dataset uses standard ontologies
including the Bibliography Ontology~\cite{Giasson2011} and the Citation Typing Ontology (CiTO)~\cite{Shotton2010} for literature
references and citations, and domain ontologies such the Protein Ontology\cite{Natale2011} and the Chemical Information
Ontology\cite{Hastings2011}. Throughout this paper various prefixes are used to denote different ontology 
namespaces to simplify the RDF examples. These prefixes are outlined in Table~\ref{namespaces}.

To expose the ChEMBL-RDF data two approaches have been adopted. First, a SPARQL end point is being
hosted at Uppsala University, using the Open-Source Edition of the Virtuoso software~\cite{Virtuoso}. Usage is free, but the
allowed volume of querying is capped, based on the estimated computational effort. Second, resources have
been made dereferencable via the \url{http://linkedchemistry.info} resource. The Kasabi
platform~\cite{kasabi} was used for a period of time, until that project was discontinued.
The Uppsala SPARQL end point still uses resource URIs based on the Kasabi approach, following a similar
URI pattern where linkedchemistry.info/chembl is replaced by data.kasabi.com/dataset/chembl-rdf.

Although many parts of the ChEMBL-RDF 13 dataset now use community based ontologies, some 
terms from the previous ad-hoc ontology are still used when a community alternative is not yet available. 
These terms were created under the 
URI namespace \textit{http://rdf.farmbio.uu.se/chembl/onto/\#}, and are referenced here using the
prefix: \textit{chembl}.

@@

We present here the updated ChEMBL-RDF, including a description of the mapping process 
and the identification of links to other datasets. Because the primary purpose of of this
paper is to expose the ChEMBL data as linked data, less focus has been placed on ontologically
capturing the fine details of the pharmacological literature. This is the reason why
many concepts are not mapped to existing ontologies, though also the lack of ontologies
with a clear open license, as defined by the Open Definition~\cite{OpenDefinition}, restrict the options.

@@

For drug discovery, the drug-like compounds themselves are the main topic of study.
ChEMBL contains many different compound types, mostly small molecules,
but also peptides, proteins, antibodies, oligosaccharides, oligonucleotides, and
even cells. ChEMBL-RDF follows the approach used by the CHEMINF ontology,
and encodes drugs as classes, rather than instances, and they subclass
other classes defined by several ontologies for these types. For example, protein drugs subclass the
Protein class from the PRotein Ontology (PR\_000000001), instead that these drugs are
of the Protein type. Each drug is a class itself. Likewise, small
molecules subclass the Chemical Entity class from the CHEMINF
ontology (CHEMINF\_000000) (which is itself equivalent to the Chemical Entity class in ChEBI, CHEBI\_24431), 
and oligosaccharides and oligonucleotides
subclass their respective matches in the CHEBI ontology (CHEBI\_50699
and CHEBI\_7754). To ensure we can look up all entities for which activities
are reported, we subclass everything from the BFO 1.1 class MaterialEntity~\cite{Smith2004},
which is a superclass of the CHEMINF Chemical Entity class.

It should also be noted that all drug-like entities in ChEMBL are not ontologically
subclassing a drug class. Instead, we give the entities the role of drug, and only within a
particular context would they be a drug. If not used as medication, they are not a drug.
Moreover, the triples only attach this drug role to approved drugs, as defined by the
ChEMBL database. 
% FIXME: What does the following exactly mean? It isn't clear whether it actually means:
% Not all drug-like entities in the database have the role of drug
All drug-like entities in the database do not have the role of drug.
The role is triplified using the OBO and ChEBI ontologies in the following manner,
where CHEBI\_23888 is the ontological entry for ``drug role'' in ChEBI:

@@

The name and synonyms for each drug are
provided as labels. If SMILES, InChI, or InChIKeys are given for a drug, then these are
provided via the CHEMINF formalism.
The CHEMINF ontology has been used here because of it's flexibility in defining
descriptor types and their origin and meaning. By allowing an ontology of descriptors,
we enable comparing values between databases semantically. Differences in, for example,
the SMILES may then be explained by the software used for that descriptor.

@@

Similarly, properties of small molecules are available from ChEMBL, and as of ChEMBL-RDF 13
these too are exposed in triple format. Like the InChI and InChIKey, these are provided using the
CHEMINF ontology. Here is, for example, the logP value (as distributed with ChEMBL and
calculated with ACD/Labs software):

@@

The literature from which all this data was extracted, is replicated from the database,
but only a subset of the possible information about that literature is provided. In all cases 
the PubMed Identifiers (PMIDs) and Digital Object Identifiers (DOIs) are used to link
to an authoritative database, as outlined later in this paper.
The properties are provided using the BIBO ontology, which is an open sourced ontology
for bibliographic information, compatible with the CiTO ontology:

@@

Each release of the ChEMBL database is accompanied by a set of release statistics, which help 
provide a concise content overview of the release in question. The ChEMBL 13 release counts
are as follows:
\begin{itemize}
  \item 8,845 targets
  \item 1,143,682 distinct molecules
  \item 6,933,068 activities
  \item 617,681 assays
  \item 44,682 documents
\end{itemize}

@@

A similar SPARQL query, replacing \verb+chembl:Target+ with \verb+bfo:MaterialEntity+, returns 1,143,682 entities for which activities are captured in ChEMBL, and by using \verb+chembl:Assay+, the query returns the total of 617,676 ChEMBL assays.
It should be noted that last query returns five entries less than the SQL database, because the
triple generation requires the assay type to be defined, which is not the case for those 
five assays. With \verb+chembl:Activity+ and \verb+bibo:Article+, we count 6,933,068 ChEMBL activities and 44,681 documents,
respectively. The document count is one fewer than appears 
in the ChEMBL 13 release notes~\cite{ChEMBL13ReleaseNotes}, as the result does not include the doc\_id -1, as this refers to the 
unpublished dataset, not itself a valid \verb+bibo:Article+.

Reproduction of the ChEMBL release statistics through querying ChEMBL-RDF with SPARQL 
provides end users with a high degree of confidence that the ChEMBL-RDF contains the same 
core content as the source relational database. The development of a validation test suite
is being discussed but not yet implemented; testing criteria other then the here presented
summarizing counts would change with each release due to curation events. Interactive,
manual curation of the conversion process as employed by the authors already is more effective
at this moment.

@@

ChEMBL-RDF can be used to explore interrelated scientific datasets, and we here present five applications,
describing how the SPARQL end point can be used, the combination of the ChEMBL data
with other life sciences databases, the use of the bibliographic information in calculating citation
statistics, and how to integrate ChEMBL-RDF with ChemSpider to provide an extension for the decision
support platform in Bioclipse.

@@

Using the SPARQL end point, it is easy to integrate data from ChEMBL into other web resources, such as CitedIn (http://www.citedin.org)~\cite{CitedIn2011}.
While the impact of a scientific publication is traditionally measured by frequency of citation in subsequent publications, this
narrowly defined calculation increasingly fails to capture the true reach of a paper. CitedIn was developed to make scientific
impact more visible and more accurate by providing insight into where published work is being cited online in a variety of public
resources, as listed on the CitedIn website~\cite{CitedResources}.
ChEMBL contains many literature references. The SPARQL endpoint is queried by CitedIn and the results are presented to its users.
Figure~\ref{fig:citedin} shows the results for the Nature paper by Gamo presenting the malaria structures~\cite{Gamo2010}
which is cited as the reference for a large number of activity datapoints and compounds in ChEMBL.
Counting all objects that reference this paper is done by the following SPARQL query template:

@@

In this study we extended Bioclipse DS with remote access to ChEMBL-RDF and ChemSpider. 
This enables users browsing chemical structures in Bioclipse to look up near neighbors in ChemSpider 
via the ChemSpider Web API (SOAP), and for the found compounds to be used to query ChEMBL-RDF for 
available interaction data. The results are presented alongside predictive models in Bioclipse
(see Fig~\ref{fig:bioclipse-ds}), and can be used for decision support when evaluating chemical
structures and considering strategies for optimization.

@@

Some information in the databases has not yet been annotated using standard ontologies. 
For example, the activity types in ChEMBL, such as ``IC50'', ``EC50'', and ``Potency'', are not exposed in the RDF dataset,
meaning that it is not yet possible to reason or group over similar activity types.
When the activity types are exposed, queries could be constructed based on the knowledge 
that the IC$_{50}$ and K$_i$ end points are closely related. Similarly, measurement units are currently provided as
string literals, making it difficult to compare concentrations when some are given in nM and others in mM.
In the future it may also be possible, relying on some assumptions, to crossmatch activities
expressed in $\mu$g/ml with query concentrations in nM. These additional semantics are necessary 
to translate raw experimental data into semantically useful, machine-readable information.

@@

The authors acknowledge A. L\"ovgren at the BMC Computing Department at Uppsala University for his
support in hosting the SPARQL end point. OS acknowledges funding from the Swedish VR (2011-6129) and eSSENCE. 
JH acknowledges funding from the EU-OPENSCREEN project. 
The data in ChEMBL is made available by funding from the Wellcome Trust [086151/Z/08/Z].

>>>>

The ChEMBL version used in this paper is ChEMBL 13, which was released on 29 February 2012.
The SQL data dump provided by ChEMBL was inserted into a local MySQL database. A set of SQL queries 
were then executed to generate RDF triples using a set of PHP scripts. The scripts use SQL to
query the database and output triples using print statements. Custom PHP scripts have been used
as a lightweight solution to provide a Linked Data API via a Apache webserver, as well as to
create data files with triples. These scripts are Open Source 
and are available from the source code hosting service, GitHub~\cite{ChEMBLRDFGitHub}. This process 
has been outlined in a best practices note by the W3C Health Care and Life Sciences Interest
Group~\cite{Marshall2012}. This process is similar to its original description.

However, the previous ChEMBL-RDF conversion used custom predicates and classes to represent RDF triples, defining
an ad hoc ontology~\cite{Willighagen2011}. This implicit ontology reflects the concepts expressed
in the relational database, and is not formally defined. The current version, instead, uses community-proposed
ontologies, making the resulting RDF triples more interoperable with RDF provided by other life sciences databases. 
The new triple dataset uses standard ontologies
including the Bibliography Ontology~\cite{Giasson2011} and the Citation Typing Ontology (CiTO)~\cite{Shotton2010} for literature
references and citations, and domain ontologies such as the Protein Ontology\cite{Natale2011} and the Chemical Information
Ontology\cite{Hastings2011}. Throughout this paper various prefixes are used to denote different ontology 
namespaces to simplify the RDF examples. These prefixes are outlined in Table~\ref{namespaces}.

To expose the ChEMBL-RDF data two approaches have been adopted. First, a SPARQL endpoint is being
hosted at Uppsala University, using the Open-Source Edition of the Virtuoso software~\cite{Virtuoso}. Usage is free, but the
allowed volume of querying is capped, based on the estimated computational effort. Second, resources have
been made dereferencable via the \url{http://linkedchemistry.info} resource. The Kasabi
platform~\cite{kasabi} was used for a period of time, until that project was discontinued.
The Uppsala SPARQL endpoint still uses resource URIs based on the Kasabi approach, following a similar
URI pattern where linkedchemistry.info/chembl is replaced by data.kasabi.com/dataset/chembl-rdf.

Although many parts of the ChEMBL-RDF 13 dataset now use community based ontologies, some 
terms from the previous ad hoc ontology are still used when a community alternative is not yet available. 
These terms were created under the 
URI namespace \textit{http://rdf.farmbio.uu.se/chembl/onto/\#}, and are referenced here using the
prefix: \textit{chembl}.

@@

We present here the updated ChEMBL-RDF, including a description of the mapping process 
and the identification of links to other datasets. Because the primary purpose of this
paper is to expose the ChEMBL data as linked data, less focus has been placed on ontologically
capturing the fine details of the pharmacological literature. This is the reason why
many concepts are not mapped to existing ontologies, though also the lack of ontologies
with a clear open license, as defined by the Open Definition~\cite{OpenDefinition}, restrict the options.

@@

For drug discovery, the drug-like compounds themselves are the main topic of study.
ChEMBL contains many different compound types, mostly small molecules,
but also peptides, proteins, antibodies, oligosaccharides, oligonucleotides, and
even cells. ChEMBL-RDF follows the approach used by the CHEMINF ontology,
and encodes drugs as classes, rather than instances, and they subclass
other classes defined by several ontologies for these types. For example, protein drugs subclass the
Protein class from the PRotein Ontology (PR\_000000001), instead that these drugs are
of the Protein type. Each drug is a class itself. Likewise, small
molecules subclass the Chemical Entity class from the CHEMINF
ontology (CHEMINF\_000000) (which is itself equivalent to the Chemical Entity class in ChEBI, CHEBI\_24431), 
and oligosaccharides and oligonucleotides
subclass their respective matches in the CHEBI ontology (CHEBI\_50699
and CHEBI\_7754). To ensure that we can look up all entities for which activities
are reported, we subclass everything from the BFO 1.1 class MaterialEntity~\cite{Smith2004},
which is a superclass of the CHEMINF Chemical Entity class.

It should also be noted that all drug-like entities in ChEMBL are not ontologically
subclassing a drug class. Instead, we give the entities the role of drug, and only within a
particular context would they be a drug. If not used as a medication, they are not a drug.
Moreover, the triples only attach this drug role to approved drugs, as defined by the
ChEMBL database. 
% FIXME: What does the following exactly mean? It isn't clear whether it actually means:
% Not all drug-like entities in the database have the role of drug
All drug-like entities in the database do not have the role of drug.
The role is triplified using the OBO and ChEBI ontologies in the following manner,
where CHEBI\_23888 is the ontological entry for ``drug role'' in ChEBI:

@@

The name and synonyms for each drug are
provided as labels. If SMILES, InChI, or InChIKeys are given for a drug, then these are
provided via the CHEMINF formalism.
The CHEMINF ontology has been used here because of its flexibility in defining
descriptor types and their origin and meaning. By allowing an ontology of descriptors,
we enable the comparison of values between databases semantically. Differences in, for example,
the SMILES string may then be explained by the software used for that descriptor.

@@

Similarly, the properties of small molecules are available from ChEMBL, and as of ChEMBL-RDF 13
these too are exposed in triple format. Like the InChI and InChIKey, these are provided using the
CHEMINF ontology. Here is, for example, the logP value (as distributed with ChEMBL and
calculated with ACD/Labs software):

@@

The literature from which all this data was extracted, is replicated from the database,
but only a subset of the possible information about that literature is provided. In all cases 
the PubMed Identifiers (PMIDs) and Digital Object Identifiers (DOIs) are used to link
to an authoritative database, as outlined later in this paper.
The properties are provided using the BIBO ontology, which is an open source ontology
for bibliographic information, compatible with the CiTO ontology:

@@

Each release of the ChEMBL database is accompanied by a set of release statistics which help 
provide a concise content overview of the release in question. The ChEMBL 13 release counts
are as follows:
\begin{itemize}
  \item 8,845 targets
  \item 1,143,682 distinct molecules
  \item 6,933,068 activities
  \item 617,681 assays
  \item 44,682 documents
\end{itemize}

@@

A similar SPARQL query, replacing \verb+chembl:Target+ with \verb+bfo:MaterialEntity+, returns 1,143,682 entities for which activities are captured in ChEMBL, and by using \verb+chembl:Assay+, the query returns the total of 617,676 ChEMBL assays.
It should be noted that this last query returns five entries less than the SQL database, because the
triple generation requires the assay type to be defined, which is not the case for those 
five assays. With \verb+chembl:Activity+ and \verb+bibo:Article+, we count 6,933,068 ChEMBL activities and 44,681 documents,
respectively. The document count is one fewer than appears 
in the ChEMBL 13 release notes~\cite{ChEMBL13ReleaseNotes}, as the result does not include the doc\_id -1, as this refers to the 
unpublished dataset, not itself a valid \verb+bibo:Article+.

Reproduction of the ChEMBL release statistics through querying ChEMBL-RDF with SPARQL 
provides end users with a high degree of confidence that the ChEMBL-RDF contains the same 
core content as the source relational database. The development of a validation test suite
is being discussed but not yet implemented; testing criteria other then the summarizing counts
presented here, would change with each release due to curation events. Interactive,
manual curation of the conversion process as employed by the authors already is more effective
at this moment.

@@

ChEMBL-RDF can be used to explore interrelated scientific datasets, and we present here five applications,
describing how the SPARQL endpoint can be used, the combination of the ChEMBL data
with other life sciences databases, the use of the bibliographic information in calculating citation
statistics, and how to integrate ChEMBL-RDF with ChemSpider to provide an extension for the decision
support platform in Bioclipse.

@@

Using the SPARQL endpoint, it is easy to integrate data from ChEMBL into other web resources, such as CitedIn (http://www.citedin.org)~\cite{CitedIn2011}.
While the impact of a scientific publication is traditionally measured by frequency of citation in subsequent publications, this
narrowly defined calculation increasingly fails to capture the true reach of a paper. CitedIn was developed to make scientific
impact more visible and more accurate by providing insight into where published work is being cited online in a variety of public
resources, as listed on the CitedIn website~\cite{CitedResources}.
ChEMBL contains many literature references. The SPARQL endpoint is queried by CitedIn and the results are presented to its users.
Figure~\ref{fig:citedin} shows the results for the Nature paper by Gamo presenting the malaria structures~\cite{Gamo2010}
which is cited as the reference for a large number of activity datapoints and compounds in ChEMBL.
Counting all objects that reference this paper is done by the following SPARQL query template:

@@

In this study we extended Bioclipse DS with remote access to both ChEMBL-RDF and ChemSpider. 
This enables users browsing chemical structures in Bioclipse to look up nearest neighbors in ChemSpider 
via the ChemSpider Web API (SOAP), and for the found compounds to be used to query ChEMBL-RDF for 
available interaction data. The results are presented alongside predictive models in Bioclipse
(see Fig~\ref{fig:bioclipse-ds}), and can be used for decision support when evaluating chemical
structures and considering strategies for optimization.

@@

Some information in the databases has not yet been annotated using standard ontologies. 
For example, the activity types in ChEMBL, such as ``IC50'', ``EC50'', and ``Potency'', are not exposed in the RDF dataset,
meaning that it is not yet possible to reason or group over similar activity types.
When the activity types are exposed, queries could be constructed based on the knowledge 
that the IC$_{50}$ and K$_i$ endpoint are closely related. Similarly, measurement units are currently provided as
string literals, making it difficult to compare concentrations when some are given in nM and others in mM.
In the future it may also be possible, relying on some assumptions, to crossmatch activities
expressed in $\mu$g/ml with query concentrations in nM. These additional semantics are necessary 
to translate raw experimental data into semantically useful, machine-readable information.

@@

The authors acknowledge A. L\"ovgren at the BMC Computing Department at Uppsala University for his
support in hosting the SPARQL endpoint. OS acknowledges funding from the Swedish VR (2011-6129) and eSSENCE. 
JH acknowledges funding from the EU-OPENSCREEN project. 
The data in ChEMBL is made available by funding from the Wellcome Trust [086151/Z/08/Z].
